---
title: "Working with noisy values"
---

```{r, include = FALSE}
knitr::opts_chunk$set(eval = FALSE, comment = "#>")
```

## What is a noisy value?

The Census Bureau protects the 2020 data with differential privacy: random
noise was added to every count before release. The number you download is
therefore an *observed* value, not the true count.

A noisy value pairs that observed number with a **posterior**: a probability
distribution over the true count, given the noise mechanism that was
actually used. Printing one shows the observed number prefixed with `~`:

```{r}
library(noisyvalue)
noisyvalue_init()

population <- get_decennial(
  geography = "county",
  variables = c(male = "P12_002N", female = "P12_026N"),
  state = "VT")

population$value
#> <noisy_vec[noisyint]: 28>
#>  [1] ~79412 ~82233 ...
```

The `value` column is a **noisy vector** (see `noisy_vec()`). Everything
else in the tibble is an ordinary column.

## Doing arithmetic

Arithmetic on noisy values combines their posteriors, so the result is
still a noisy value you can summarize. Dependencies between values (shared
noise) are tracked, so sums and differences stay correct.

```{r}
chittenden <- population[population$GEOID == "50007", ]
orange     <- population[population$GEOID == "50017", ]

male_total <- chittenden$value[chittenden$variable == "male"] +
  orange$value[orange$variable == "male"]
```

Supported: `+ - * / ^ %% %/%`, `sum()`, `mean()`, `any()`, `all()`, `&`,
`|`, `!`. Functions with no meaningful posterior (`median()`, `sort()`,
`unique()`, ...) raise an error instead of silently using the observed
numbers.

## Summarizing a value

```{r}
noisy_credible_interval(male_total)            # central 95% interval
noisy_credible_interval(male_total, p = 0.9)
```

The interval covers where the *true* count plausibly lies. It is not the
sampling interval you would get from a survey.

## Comparing values

Comparing two single noisy values gives a noisy boolean, which has a
probability rather than a definite answer:

```{r}
male   <- chittenden$value[chittenden$variable == "male"]
female <- chittenden$value[chittenden$variable == "female"]

noisy_prob(male > female)   # P(true male count > true female count)
```

Comparing two longer columns instead gives an ordinary logical vector over
the *observed* numbers, which is what you want for filtering rows:

```{r}
population[population$value > 50000, ]
```

## Single values and columns

* `x[[i]]` returns one value; `x[i]` returns a length-one column.
* `noisy_credible_interval()` and `noisy_prob()` accept either, as long as
  it is a single value. For a whole column, reduce it first with `sum()` or
  `mean()`.
* `c()` combines noisy vectors of the same type.

## Randomness

`noisy_credible_interval()` and `noisy_prob()` estimate by simulation. Use
`n` for more draws (more precise, slower) and `rng` to fix the seed:

```{r}
noisy_prob(male > female, n = 10000, rng = 1)
```

## Finding data

```{r}
dhc_queries()               # what get_dhc() can return
get_dhc("county", "sex*hispanic", state = "VT", county = "007")
```

`get_decennial()` accepts the same variable codes as
`tidycensus::get_decennial()` (table `P12` so far) and returns the same
shape of result.
